---
title: "Data on Solitary Confinement at the Northwest Detention Center"
author: "UWCHR"
date: "2024-03-14"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

# Data source

This dataset of ICE Segregation Review Management System (SRMS) records from September 1, 2018 through September 1, 2018 was released pursuant to a FOIA request by Harvard Law School and published as [replication data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/AT7YFA) for the report ['“Endless Nightmare” Torture and Inhuman Treatment in Solitary Confinement in U.S. Immigration Detention'](https://phr.org/our-work/resources/endless-nightmare-solitary-confinement-in-us-immigration-detention) by Harvard Immigration and Refugee Clinical Program (HIRCP) and Arevik Avedian at Harvard Law School (HLS), Peeler Immigration Lab (PIL) at Harvard Medical School (HMS), and Physicians for Human Rights (PHR).

## Notebook goal

This notebook serves as an appendix to UWCHR's [NWDC Conditions Research Update: Charles Leo Daniel’s Death at NWDC in Context](https://jsis.washington.edu/humanrights/2024/03/15/nwdc-conditions-research-update-daniel-death-in-context/). Analysis focuses on ICE Segregation Review Management System (SRMS) records reported by the Northwest ICE Processing Center in Tacoma, WA; also known as the Northwest Detention Center (NWDC), and coded as "TACOMA ICE PROCESSING CENTER (NORTHWEST DET CTR) (WA)" in the below data. Given that this analysis was conducted in response to a breaking news event, it may be updated to reflect ongoing analysis.

```{r import_data, message = FALSE, warning = FALSE}

library(pacman)
p_load(tidyverse, arrow, lubridate, zoo, digest, ggplot2, plotly, gghighlight, viridis, gridExtra)

df <- read_delim(here::here('analyze', 'input', 'SRMS_spreadsheet_9.1.2018_-_9.1.2023_Redacted.csv.gz'), delim='|',
                      col_types = cols('Placement Date' = col_date(format="%m/%d/%Y"),
                                       'Release Date' = col_date(format="%m/%d/%Y"))) %>% 
  drop_na('Tracking Number', 'Facility AOR')

names(df) %<>% stringr::str_replace_all("\\s","_") %>% tolower

nwdc_det_trinidad <- read_delim(here::here('analyze', 'frozen', 'nwdc_trinidad_and_tobago_fy11-24ytd.csv.gz'), delim = "|",
                            col_types = cols(stay_book_in_date_time = col_datetime(format=''),
                                             stay_book_out_date_time = col_datetime(format=''),
                                             detention_book_in_date_and_time = col_datetime(format=''),
                                             detention_book_out_date_time = col_datetime(format='')
                                             ))

nwdc <- "TACOMA ICE PROCESSING CENTER (NORTHWEST DET CTR) (WA)"

# Checking for any variation in SEA AOR facility names; NWDC should be only one present
stopifnot(length(unique(df[df$facility_aor == "SEA (Seattle)", "facility"])) == 1)

# Data cleaning per `dataverse_files/analysis/data_prep.do`
df <- df %>% 
  mutate(mental_illness = case_when(mental_illness == "Yes" ~ "Mental Illness",
                                    TRUE ~ mental_illness),
         facility = case_when(facility == "WEBB COUNTY DETENTION CENTER (TX)" ~ "WEBB COUNTY DETENTION CENTER (CCA) (TX)",
                              TRUE ~ facility))

# Physicians for Human Rights/Harvard Law School report omits records with ongoing solitary placements
# Uncomment this filter to reproduce PHR/HLS report results:
# df <- df %>%
# filter(!is.na(release_date))

```

# Data overview

The HLS/PHR dataset includes a total of `r nrow(df)` SMRS placement records for `r length(unique(df$facility))` U.S. ICE detention facilities. For a full description of the dataset, see the [full report](https://phr.org/our-work/resources/endless-nightmare-solitary-confinement-in-us-immigration-detention).

The dataset includes a total of `r nrow(df %>% filter(facility == nwdc))` SRMS placement records for the Northwest Detention Center (NWDC) in Tacoma, WA; including `r nrow(df %>% filter(facility == nwdc, is.na(release_date)))` solitary placements that were ongoing at the time of release of the dataset. These ongoing solitary placements are included in this analysis.

## Death of Charles Leo Daniel at NWDC

On March 7, 2024, the death of Charles Leo Daniel was reported at Northwest Detention Center (NWDC). Daniel, a citizen of Trinidad and Tobago, was reportedly in segregation at the time of his death. ICE detention records processed in a separate repository show that Daniel was booked in to the NWDC on April 2, 2020, and that he was the only citizen of Trinidad and Tobago to be detained at the facility from his arrival in 2020 until August 15, 2022 (see `analyze/frozen/nwdc_trinidad_and_tobago_fy11-24ytd.csv.gz`). 

Therefore, we conclude that the two ICE SRMS records in the dataset for a male citizen of Trinidad and Tobago at NWDC represent Daniel's segregation history while detained at the facility. The full records are presented below (note redaction of some values):

```{r nwdc_seg_trinidad, message = FALSE, warning = FALSE}

nwdc_seg_trin <- df %>% 
  filter(str_detect(facility, "TACOMA"),
         country_of_citizenship == 'TRINIDAD AND TOBAGO') %>% 
  mutate(placement_date = as.character(placement_date),
         release_date = as.character(release_date))

transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  t_df <- t_df %>%
    tibble::rownames_to_column(.data = .) %>%
    tibble::as_tibble(.)
  return(t_df)
}

knitr::kable(transpose_df(nwdc_seg_trin) %>% 
               rename(column = rowname,
                      placement_1 = "1",
                      placement_2 = "2"),
             caption = "NWDC SRMS records for citizens of Trinindad and Tobago")

```

The first solitary placement began on `r nwdc_seg_trin$placement_date[1]`, shortly after Daniel was booked into the NWDC and lasted until `r nwdc_seg_trin$release_date[1]`, or `r nwdc_seg_trin$length_of_stay[1]` days. The second solitary placement began two days later on `r nwdc_seg_trin$placement_date[2]` and was ongoing at the time of release of the HLS/PHR data in mid-September 2023, or at least `r nwdc_seg_trin$length_of_stay[2]` days. (As noted in the replication data for the HLS/PHR report, ongoing solitary placements were dropped from analysis for the published report; we include these records here due to their relevance to issues at the NWDC.)

As noted, both stays were reported by ICE as "Detainee Request", with the second stay categorized as "Protective Custody". Both stays are also flagged as "Serious Mental Illness".

As seen below, Daniel's second placement of at least `r nwdc_seg_trin$length_of_stay[2]` days was the second longest individual solitary placement reported nationally during the time period covered by the HLS/PHR dataset. As seen below, the longest individual placement reported nationally was also at NWDC and was also ongoing at the time of release of the dataset. (Because unique personal identifiers are not included in this dataset, it is not possible to calculate cumulative stay length per person across individual solitary placements.)

```{r placemnent_length_natl, message = FALSE, warning = FALSE}

t1 <- df %>% 
  mutate(ongoing_stay = is.na(release_date)) %>% 
  arrange(desc(length_of_stay)) %>% 
  head(10) %>% 
  mutate(length_of_stay = case_when(ongoing_stay == TRUE ~ paste0(as.character(length_of_stay), "+"),
                                    TRUE ~ as.character(length_of_stay))) %>% 
  dplyr::select(country_of_citizenship, facility, length_of_stay, mental_illness) %>% 
  knitr::kable(caption = "Top 10 longest solitary placements nationwide, Sept. 2018-Sept. 2023",
               row.names = TRUE)

t1

```

As reported by Physicians for Human Rights/Harvard Law School, Northwest Detention Center ranked 9th out of `r length(unique(df$facility))` facilities represented in the SRMS data when ranked by average solitary placement length, this figure does not change when ongoing stays are taken into account; however, note below that the number of reported solitary placements at NWDC was significantly greater than other facilities in the top 10:

```{r placemnent_length_rank_natl, message = FALSE, warning = FALSE}

t0 <- df %>% 
  group_by(facility) %>% 
  summarize(avg_placement_length = mean(length_of_stay),
            n = n()) %>% 
  arrange(desc(avg_placement_length)) %>%
  head(10) %>%
  dplyr::select(facility, avg_placement_length, n) %>% 
  knitr::kable(caption = "Avg. solitary placement length by facility, Sept. 2018-Sept. 2023",
               row.names = TRUE,
               digits=2)

t0

```

## Comparison of solitary placement lengths

In fact, solitary placements reported to SRMS at the Northwest ICE Process Center were significantly longer on average than nationally, with an average stay length of `r round(mean(df[df$facility == nwdc,]$length_of_stay), 2)` days compared to a national average of `r round(mean(df$length_of_stay), 2)` days, and a median stay length of `r round(median(df[df$facility == nwdc,]$length_of_stay), 2)` days compared to a national median of `r round(median(df$length_of_stay), 2)` days:

```{r compare_nwdc_vs_all, message = FALSE, warning = FALSE}

df_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  mutate(group = "NWDC")

df_all <- df %>% 
  mutate(group = "Nat'l")

dat <- rbind(df_nwdc, df_all) %>% 
  mutate(group = as.factor(group))

b1 <- dat %>% 
  group_by(group) %>% 
  ggplot(aes(x = group, y = log(length_of_stay), fill=group)) +
  geom_boxplot() +
  scale_fill_viridis_d(direction=-1, breaks=rev) +
  scale_x_discrete() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       subtitle = "National versus NWDC",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1

```

A one-sample t-test shows that this difference in mean stay lengths is statistically significant:

```{r one_sample_t_test}

# Data are approximately normally distributed; we expect banding around stay lengths associated with ICE reporting requirements i.e. 3, 14, 30 days
# p1 <- qqnorm(log(df$length_of_stay))
# p2 <- qqnorm(log(df[df$facility == nwdc,]$length_of_stay))

dat <- log(df[df$facility == nwdc,]$length_of_stay)
mu <- mean(log(df$length_of_stay))

t.test(x=dat, mu=mu, alternative = "greater")

```

## Special vulnerabilities

SRMS placements are flagged for "special vulnerability" categories, noted in the dataset as "Mental Illness", "Serious Medical Illness", "Serious Disability", and "Suicide Risk?". The following charts and plots summarize the proportion of SRMS records flagged for these categories at NWDC versus nationally; as well as distributions of placement length for each category at NWDC versus nationally. We note that in almost all categories, solitary placements at NWDC tend to be longer than for the same categories nationwide.

### Mental illness

```{r mental_illness_prop, message=FALSE, warning=FALSE}

df$mental_illness <- factor(df$mental_illness, levels = c("Serious Mental Illness", "Mental Illness", "None", NA))

mi_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  count(mental_illness) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

mi_natl <- df %>%
  count(mental_illness) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

dat <- left_join(mi_natl, mi_nwdc, by='mental_illness', suffix=c(".natl", ".nwdc")) %>% 
  pivot_longer(cols=-c('mental_illness')) %>%
  filter(str_detect(name, "prop"))

p1 <- dat %>% 
  ggplot(aes(x = mental_illness, y = value, fill=name), color = 'black') +
    geom_col(position='dodge') +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d(limits=rev) +
  labs(title = "Proportion of solitary placements",
       subtitle = "NWDC vs. national",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS") +
  theme_minimal()

p1


```

```{r mental_illness_length, message=FALSE, warning=FALSE}

df_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  mutate(group = "NWDC")

df_all <- df %>% 
  mutate(group = "National")

dat <- rbind(df_nwdc, df_all) %>% 
  mutate(group = as.factor(group),
         mental_illness = case_when(mental_illness %in% c("None", NA) ~ "None/NA",
         TRUE ~ mental_illness),
         mental_illness = factor(mental_illness, levels = c("Serious Mental Illness", "Mental Illness", "None/NA")))

b1 <- dat %>% 
  ggplot(aes(x = group, y = log(length_of_stay), fill = mental_illness)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       subtitle = "National versus NWDC",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1

```

```{r mental_illness_length_tab, message=FALSE, warning=FALSE}

dat %>% 
  group_by(group, mental_illness) %>% 
  summarize("average placement length" = mean(length_of_stay, na.rm=TRUE),
            "median placement length" = median(length_of_stay, na.rm=TRUE),
            "total placements" = n()) %>% 
  pivot_wider(names_from = c('group'), names_glue = "{group} {.value}", values_from = c('average placement length', 'median placement length', "total placements")) %>% 
  mutate("NWDC % longer than national average" = (`NWDC average placement length` / `National average placement length` - 1) * 100,
         "NWDC % longer than national median" = (`NWDC median placement length` / `National median placement length` - 1) * 100
         ) %>% 
  # dplyr::select(-c("NWDC % longer than national average",
  #                   "NWDC % longer than national median")) %>% 
  knitr::kable(digits=2)



```

```{r mental_illness_placements_fig, message=FALSE, warning=FALSE}

p1 <- df %>% 
  filter(facility == nwdc,
         placement_date >= "2018-09-01") %>% 
  mutate(qtr = as.factor(quarter(placement_date, fiscal_start=10, type="year.quarter"))) %>% 
  count(qtr, mental_illness) %>% 
  ggplot(aes(x = qtr, y = n, fill = mental_illness)) +
  geom_col(position='stack') +
  labs(title="NWDC solitary placements, Sept. 2018-Sept. 2023",
       subtitle="`Mental Illness` categories",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS") +
  ylab("Total Placements") +
  xlab("Placement date (FY quarter)") +
  scale_fill_viridis_d(na.value='grey') +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1))
  
p1

```

### Serious medical illness

```{r medical_illness_length, message=FALSE, warning=FALSE}

df_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  mutate(group = "NWDC")

df_all <- df %>% 
  mutate(group = "Nat'l")

dat <- rbind(df_nwdc, df_all) %>% 
  mutate(group = as.factor(group))

# dat %>% 
#   count(group, serious_medical_illness)

b1 <- dat %>% 
  ggplot(aes(x = group, y = log(length_of_stay), fill = serious_medical_illness) )+
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       subtitle = "National versus NWDC",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1

```

### Serious disability

```{r serious_disability_length, message=FALSE, warning=FALSE}

df_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  mutate(group = "NWDC")

df_all <- df %>% 
  mutate(group = "Nat'l")

dat <- rbind(df_nwdc, df_all) %>% 
  mutate(group = as.factor(group))

# dat %>% 
#   count(group, serious_disability)

b1 <- dat %>% 
  ggplot(aes(x = group, y = log(length_of_stay), fill = serious_disability) )+
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       subtitle = "National versus NWDC",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1

```

## Detainee-requested versus Facility-initiated solitary placements

Note larger proportion of "detainee-requested" solitary placements reported at NWDC compared to national; similar distribution of placement lengths between these categories comparing NWDC to national stay length.

```{r detainee_request_prop, message=FALSE, warning=FALSE}

dr_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  count(detainee_request_segregation) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

dr_natl <- df %>%
  count(detainee_request_segregation) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

dat <- left_join(dr_natl, dr_nwdc, by='detainee_request_segregation', suffix=c(".natl", ".nwdc")) %>% 
  pivot_longer(cols=-c('detainee_request_segregation')) %>%
  filter(str_detect(name, "prop"))

p1 <- dat %>% 
  ggplot(aes(x = detainee_request_segregation, y = value, fill=name), color = 'black') +
    geom_col(position='dodge') +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d(limits=rev) +
  labs(title = "Proportion of solitary placements",
       subtitle = "NWDC vs. national",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS") +
  theme_minimal()

p1

```

```{r detainee_request_length, message=FALSE, warning=FALSE}

df_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  mutate(group = "NWDC")

df_all <- df %>% 
  mutate(group = "Nat'l")

dat <- rbind(df_nwdc, df_all) %>% 
  mutate(group = as.factor(group))

b1 <- dat %>% 
  ggplot(aes(x = group, y = log(length_of_stay), fill = detainee_request_segregation)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       subtitle = "National versus NWDC",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1

```

## Placement reason

Note no placements categorized as "Hunger Strike", "Mental Illness" (some "Medical/Mental"), or "Suicide Risk" at NWDC. Note several of these categories are not mutually-exclusive.

```{r placement_reason_length, message=FALSE, warning=FALSE}

df <- df %>% 
  mutate(placement_reason_short = case_when(
    str_detect(placement_reason, "Detainee Request: Other") ~ "Detainee Request",
    str_detect(placement_reason, "Disciplinary") ~ "Disciplinary",
    str_detect(placement_reason, "Facility Initiated: Other") ~ "Facility Initiated",
    str_detect(placement_reason, "Facility Security Threat") ~ "Facility Security Threat",
    str_detect(placement_reason, "Facility Security Threat: Due to Seriousness of Criminal Conviction") ~ "Facility Security Threat",
    str_detect(placement_reason, "Facility Security Threat: Gang Member Status (Not Protective Custody)") ~ "Facility Security Threat",
    str_detect(placement_reason, "Facility Security Threat: Other") ~ "Facility Security Threat",
    str_detect(placement_reason, "Facility Security Threat: Violent or Disruptive Behavior") ~ "Facility Security Threat",
    str_detect(placement_reason, "Hunger Strike") ~ "Hunger Strike",
    str_detect(placement_reason, "Medical/Mental") ~ "Medical/Mental",
    str_detect(placement_reason, "Medical: Detox/Withdrawal Observation") ~ "Medical",
    str_detect(placement_reason, "Medical: Disabled or Infirm") ~ "Medical",
    str_detect(placement_reason, "Medical: Observation") ~ "Medical",
    str_detect(placement_reason, "Medical: Other Infectious Disease") ~ "Medical",
    str_detect(placement_reason, "Medical: Other") ~ "Medical",
    str_detect(placement_reason, "Medical: Segregation Unit") ~ "Medical",
    str_detect(placement_reason, "Medical: Tuberculosis") ~ "Medical",
    str_detect(placement_reason, "Mental Illness") ~ "Mental Illness",
    str_detect(placement_reason, "Mental Illness: Observation") ~ "Mental Illness",
    str_detect(placement_reason, "Other: Pending General Population Bedspace Availability") ~ "Other",
    str_detect(placement_reason, "Other: Unique Reason Not Listed (Explain in Comments)") ~ "Other",
    str_detect(placement_reason, "Pending Investigation of Disciplinary Violation") ~ "Pending Disciplinary",
    str_detect(placement_reason, "Protective Custody") ~ "Protective Custody",
    str_detect(placement_reason, "Protective Custody: Criminal Offense (i.e. Sex Offender)") ~ "Protective Custody",
    str_detect(placement_reason, "Protective Custody: Gang Status (Protective Custody Only)") ~ "Protective Custody",
    str_detect(placement_reason, "Protective Custody: Lesbian, Gay, Bisexual, Transgender (LGBT)") ~ "Protective Custody",
    str_detect(placement_reason, "Protective Custody: Other Detainee Safety") ~ "Protective Custody",
    str_detect(placement_reason, "Protective Custody: Special Vulnerability Other") ~ "Protective Custody",
    str_detect(placement_reason, "Protective Custody: Victim of Sexual Assault") ~ "Protective Custody",
    str_detect(placement_reason, "Suicide Risk Placement") ~ "Suicide Risk",
    str_detect(placement_reason, "Other") ~ "Other",
  ))

df_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  mutate(group = "NWDC")

df_all <- df %>% 
  mutate(group = "Nat'l")

dat <- rbind(df_nwdc, df_all)

b1 <- dat %>% 
  ggplot(aes(y = placement_reason_short, x = log(length_of_stay), fill = group)) +
  geom_boxplot() +
  scale_fill_viridis_d(limits=rev) +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       subtitle = "National versus NWDC",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1
```

```{r placement_reason_prop, message=FALSE, warning=FALSE}

dr_nwdc <- df %>% 
  filter(facility == nwdc) %>% 
  count(placement_reason_short) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

dr_natl <- df %>%
  count(placement_reason_short) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

dat <- left_join(dr_natl, dr_nwdc, by='placement_reason_short', suffix=c(".natl", ".nwdc")) %>% 
  pivot_longer(cols=-c('placement_reason_short')) %>%
  filter(str_detect(name, "prop"))

p1 <- dat %>% 
  ggplot(aes(y = placement_reason_short, x = value, fill=name), color = 'black') +
    geom_col(position='dodge') +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_viridis_d(limits=rev) +
  labs(title = "Proportion of solitary placements",
       subtitle = "NWDC vs. national",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS") +
  theme_minimal()

p1

```

## Attorney representation

Attorney representation is associated with statistically significant difference in stay length.

Avg. stay w/ attorney representation is `r round(mean(df[df$attorney_of_record == "Yes",]$length_of_stay), 2)` days; avg. stay w/o attorney representation: `r round(mean(df[df$attorney_of_record == "No/NA",]$length_of_stay), 2)`.

```{r attorney_rep_length_dist}

df <- df %>% 
  mutate(attorney_of_record = case_when(attorney_of_record %in% c("No", NA) ~ "No/NA",
         TRUE ~ attorney_of_record))

b1 <- df %>% 
  ggplot(aes(x = attorney_of_record, y = log(length_of_stay), fill = attorney_of_record)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1


dat <- log(df[df$attorney_of_record == "Yes",]$length_of_stay)
mu <- mean(log(df$length_of_stay))

t.test(x=dat, mu=mu, alternative = "less")



```

When selecting only records with attorney representation, attorney notification is not associated with statistically-significant difference in stay length:

```{r attorney_rep_length_dist}

b1 <- df %>% 
  filter(attorney_of_record == "Yes") %>% 
  mutate(attorney_notification = case_when(attorney_notification %in% c("No", NA) ~ "No/NA",
         TRUE ~ attorney_notification)) %>% 
  ggplot(aes(x = attorney_notification, y = log(length_of_stay), fill = attorney_notification)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Distribution of solitary placement lengths",
       caption="Figure: UWCHR analysis\nData source: ICE SRMS data released via FOIA to PHR/HLS")

b1

dat <- df %>% 
  filter(attorney_of_record == "Yes") %>% 
  mutate(attorney_notification = case_when(attorney_notification %in% c("No", NA) ~ "No/NA",
         TRUE ~ attorney_notification))

attorney_notified <- log(dat[dat$attorney_notification == "Yes",]$length_of_stay)
mu <- mean(log(dat$length_of_stay))

t.test(x=attorney_notified, mu=mu, alternative = "two.sided")

```


## Interactive charts

The following interactive charts offer an overview of detailed "Placement Reason" values over time at NWDC and nationally:

### NWDC solitary placements by "Placement Reason"

```{r nwdc_placement_reason_interactive, message=FALSE, warning=FALSE}

p1 <- df %>% 
  filter(facility == nwdc,
         placement_date >= "2018-09-01") %>% 
  mutate(qtr = as.factor(quarter(placement_date, fiscal_start=10, type="year.quarter")),
         placement_reason = str_wrap(placement_reason, width=25)) %>% 
  count(qtr, placement_reason) %>% 
  ggplot(aes(x = qtr, y = n, fill = placement_reason)) +
  geom_col(position='stack') +
  labs(title="NWDC solitary placements, Sept. 2018-Sept. 2023",
       subtitle="`Placement Reason` categories",
       caption="Data source: ICE SRMS data released via FOIA to PHR/HLS\nFigure: UWCHR analysis") +
  ylab("Placements") +
  xlab("Placement date (FY quarter)") +
  scale_fill_viridis_d(na.value='grey') +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1))
  
ggplotly(p1)

```

### National solitary placements by "Placement Reason"

```{r natl_placement_reason_interactive, message=FALSE, warning=FALSE}

p1 <- df %>% 
  filter(placement_date >= "2018-09-01") %>% 
  mutate(qtr = as.factor(quarter(placement_date, fiscal_start=10, type="year.quarter")),
         placement_reason = str_wrap(placement_reason, width=25)) %>% 
  count(qtr, placement_reason) %>% 
  ggplot(aes(x = qtr, y = n, fill = placement_reason)) +
  geom_col(position='stack') +
  labs(title="National ICE solitary placements, Sept. 2018-Sept. 2023",
       subtitle="`Placement Reason` categories",
       caption="Data source: ICE SRMS data released via FOIA to PHR/HLS\nFigure: UWCHR analysis") +
  ylab("Placements") +
  xlab("Placement date (FY quarter)") +
  scale_fill_viridis_d(na.value='grey') +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1))
  
ggplotly(p1)

```
